Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-10 manuscript/manuscript.Rmd
updates sensitivity results

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.

@@

### Vital rate sensitivities to climate (IPM)

### Forecasting Climate Change Impacts
Equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Figure 3). For *B. gracilis*, the IPM predicted a modest increase in cover with a 1% increase in the mean of precipitation or a 1% increase in the mean of temperature, the compounding effect of both being a 20% increase in cover. This reflects the relatively strong effects of precipitation and climate on *B. gracilis* genet growth and recruitment (Figure 1). The QBM also predicted increased *B. gracilis* cover with a precipitation increase, but increasing temperature decreased equilibrium cover (Figure 3).

@@

It is perhaps startling that two different types of models fit using the same data produce drastically different forecasts (Figure 3).

@@

@@

Figures
-------
```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.width=5, fig.height=8}
library(png)
library(grid)
img <- readPNG("components/figure/micromeso_flow.png")
grid.raster(img)
```

@@

mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
                    value = mean(eq_cover))

myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
ggplot(equilibrium_cover, aes(x=sim, y=eq_cover*100, 
                              linetype=vital, group=vital, shape=vital))+
  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
  geom_line()+
  geom_point(size=4)+
  facet_grid(species~., scales = "free")+
  scale_linetype_manual(values=c(1,2,3), name="Vital Rate")+
  scale_shape_manual(values=c(19,17,15), name="Vital Rate")+
  ylab("Equilibrium Cover (%)")+
  xlab("Simulation")+
  theme_bw()
setwd("~/Repos/MicroMesoForecast/manuscript")
```

```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM. Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure. Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_percdiffs.rds")
qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_percdiffs.rds")
ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))

@@

myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
# ggplot(climSimsQBM)+
#   geom_boxplot(aes(x=species, y=cover*100, fill=sim), outlier.size=0)+
#   theme_bw()+
#   coord_cartesian(ylim=c(0,50))+
#   scale_fill_manual(name="", labels=c("Observed", "1% Precipitation Increase", "1% Temperature Increase", "1% Temp. & Ppt. Increase"), values=myCols2)
ggplot(diff_df, aes(x=sim, y=value, fill=type))+
  geom_bar(stat="identity", position="dodge")+
  geom_hline(aes(yintercept=0))+
  scale_fill_manual(values = myCols2[c(1,3)], name="Model type")+
  xlab("Climatic change")+
  ylab("Proportional change in percent cover")+
  facet_grid(.~variable)+
  scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
  theme_bw()
```

>>>>

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit a single regression model for population growth. We describe the statistical models separately since fitting the models required different approaches. All models contain five climate covariate that we chose \emph{a priori}: "water year" precipitation at \emph{t}-1 (lagppt); fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year. We also include interactions among same-year climate covariates (e.g., ppt1 $\times$ TmeansSpr1) and climate $\times$ size interactions.

@@

### Forecasting Climate Change Impacts
Simulated equilibrium cover simulated from the models was sensitive to climate perturbations, but the IPM and QBM produced inconsistent results (Table 2). For the IPM, proportional changes range from about 1-25% for *B. gracilis*, 7-25% for *H. comata*, 8% for *P. smithii*, and 2-20% for *P. secunda*. For the QBM, proportional changes due to climate perturbation ranged from 20-100% for *B. gracilis*, 10-14% for *H. comata*, 0.25-8% for *P. smithii*, and 4-17% for *P. secunda*. However, the forecasts were extremely uncertain (Figure 1). The IPM produced more certain forecasts than the QBM, as expected based on the validation results (Table 1). But even for the IPM forecasts, predicted percent changes in median equilibrium cover ranged from -200% to 200% (Figure 1). Forecast uncertainty did not vary appreciably among species.

The population responses of species as forecast by the IPM is a function of the combined effects of individual vital rate sensitivities to climate and the contribution of each vital to population dynamics. For *B. gracilis*, all vital rates (survival, growth, and rectruitment) were positively affected by increases in precipitation and temperature (Figure 2). Intuitively, this leads to an overall increase in *B. gracilis* cover (Table 2).

Precipitation increase resulted in all vital rates contributing to an increase in simulated cover for *H. comata*. But there is a small negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK SIMULATIONS.

*P smithii* responded negatively to an increase in precipitation, temperature, or both regardless of the vital rate through which perturbed climate was acting. The effects were small, however, leading to relatively minor population consequences under climate change that did not vary much across scenarios (Table 2).

For *B. gracilis*, the IPM predicted a modest increase in cover with a 1% increase in the mean of precipitation or a 1% increase in the mean of temperature, the compounding effect of both being a 20% increase in cover. This reflects the relatively strong effects of precipitation and climate on *B. gracilis* genet growth and recruitment (Figure 1). The QBM also predicted increased *B. gracilis* cover with a precipitation increase, but increasing temperature decreased equilibrium cover (Figure 3).

@@

### Forecasting the future, and the future of forecasting
Our goal was not make any explicit forecast for the future state of these populations based on predicted climate change. But our results highlight the state of affairs in ecology when it comes to forecasting the impacts of climate change. The analysis we conducted here could be considered, with some exceptions of course, at the forefront of ecological forecasting in terms of the statistical approach employed (hierarchical Bayesian), the type of population model we used (stochastic IPM with parameter uncertainty), and the amount of data we had at our disposal (14 years of individual-level data). Yet, model predictions proved so uncertain that any forecast, when bounded with uncertainty, would be at best not useful and at worst meaningless.

Something about fitting the models...then cite Britta's paper: 20-25 years needed!

@@

\pagebreak{}
```{r table_2,results='asis',message=FALSE, comment=NA, echo=FALSE}
qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))

diff_df <- rbind(qbm_diffs, ipm_diffs)
med_cover_perc <- as.data.frame(diff_df$med_cover*100)
med_cover_perc$model <- diff_df$type 
med_cover_perc$sim <- diff_df$variable
med_cover_perc$species <- diff_df$species
colnames(med_cover_perc)[1] <- "cover"
medcast <- dcast(med_cover_perc, species+sim~model, value.var = "cover")
colnames(medcast) <- c("Species", "Simulation", "IPM Change (%)", "QBM Change (%)")
simnames <- c("ppt", "temp", "ppt+temp")
medcast$Simulation <- rep(simnames, 4)
print.xtable(xtable(medcast, caption = "Percent change from equilibrium (temporal median of simulated plant cover) at observed climate relative to a 1\\% climate change as forecast by each model type."),type="latex", comment=FALSE,
             include.rownames=FALSE, caption.placement="top")
```

@@

Figures
-------
```{r fig.cap="Work flow of the data aggregation, model fitting, and population simulating.", fig.height=6}
library(png)
library(grid)
img <- readPNG("components/figure/micromeso_flow.png")
grid.raster(img)
```

@@

mean_cover <- ddply(subset(equilibrium_cover, sim=="allClim"), .(species), summarise,
                    value = median(eq_cover))

myCols2 <- c("#277BA8", "#7ABBBD", "#AED77A")
ggplot(equilibrium_cover, aes(x=sim, y=eq_cover*100, 
                              linetype=vital, group=vital, shape=vital))+
  geom_hline(data=mean_cover, aes(yintercept=value*100), linetype=2, color="grey45")+
  geom_line()+
  geom_point(size=4)+
  facet_grid(species~., scales = "free")+
  scale_linetype_manual(values=c(1,2,3), name="Vital Rate")+
  scale_shape_manual(values=c(19,17,15), name="Vital Rate")+
  ylab("Equilibrium Cover (%)")+
  xlab("Simulation")+
  scale_x_discrete(labels=c("obs", "ppt", "temp", "ppt + tmp"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
setwd("~/Repos/MicroMesoForecast/manuscript")
```

```{r figure_2, dependson="plot-options", fig.cap="Proportional change in species' mean cover caused by a 1% increase in observed precipitation, temperature, or both as predicted by the individual-based IPM and the aggregate-based QBM. Note that the change for POSE due to a precipitation increase predicted by the QBM is almost zero and so does not show up in the figure. Labels on x-axis refer to: 'ppt' = 1% increase in mean precipitation; 'temp' = 1% increase in mean temperature; 'ppt + temp' = 1% increase in mean precipitation and mean temperature.", fig.width=8.5, fig.height=3, cache=FALSE, results='hide'}
qbm_diffs <- readRDS("../montana/quadBM/simulations/results/qbm_climatesims_logdiffs.RDS")
qbm_diffs$type <- rep("QBM", nrow(qbm_diffs))
ipm_diffs <- readRDS("../montana/ipm/simulations/results/ipm_climatesims_logdiffs.RDS")
ipm_diffs$type <- rep("IPM", nrow(ipm_diffs))

@@

myCols2 <- c("grey45", "#277BA8", "#7ABBBD", "#AED77A")
dgd = position_dodge(width = 0.60)
ggplot(diff_df)+
  geom_hline(aes(yintercept=0), linetype=2)+
  geom_errorbar(aes(x=variable, ymin=lo_cover, ymax=up_cover, group=type),
                position=dgd, width=0.25, color="grey45")+
  geom_point(aes(x=variable, y=med_cover, shape=type),
             position=dgd, size=5, color="white")+
  geom_point(aes(x=variable, y=med_cover, shape=type),
             position=dgd, size=3)+
  scale_shape_manual(values = c(19,17), name="Model type")+
  xlab("Climatic change")+
  ylab("Proportional change in percent cover")+
  facet_grid(.~species)+
  scale_x_discrete(labels=c("ppt", "temp", "ppt + tmp"))+
  theme_bw()
```

### 2015-06-10 manuscript/manuscript.tex
updates sensitivity results

\subsubsection{Vital rate sensitivities to climate
(IPM)}\label{vital-rate-sensitivities-to-climate-ipm}

@@

Equilibrium cover simulated from the models was sensitive to climate
perturbations, but the IPM and QBM produced inconsistent results (Figure
3). For \emph{B. gracilis}, the IPM predicted a modest increase in cover
with a 1\% increase in the mean of precipitation or a 1\% increase in
the mean of temperature, the compounding effect of both being a 20\%
increase in cover. This reflects the relatively strong effects of
precipitation and climate on \emph{B. gracilis} genet growth and
recruitment (Figure 1). The QBM also predicted increased \emph{B.
gracilis} cover with a precipitation increase, but increasing
temperature decreased equilibrium cover (Figure 3).

@@

It is perhaps startling that two different types of models fit using the
same data produce drastically different forecasts (Figure 3).

>>>>

@@

Simulated equilibrium cover simulated from the models was sensitive to
climate perturbations, but the IPM and QBM produced inconsistent results
(Table 2). For the IPM, proportional changes range from about 1-25\% for
\emph{B. gracilis}, 7-25\% for \emph{H. comata}, 8\% for \emph{P.
smithii}, and 2-20\% for \emph{P. secunda}. For the QBM, proportional
changes due to climate perturbation ranged from 20-100\% for \emph{B.
gracilis}, 10-14\% for \emph{H. comata}, 0.25-8\% for \emph{P. smithii},
and 4-17\% for \emph{P. secunda}. However, the forecasts were extremely
uncertain (Figure 1). The IPM produced more certain forecasts than the
QBM, as expected based on the validation results (Table 1). But even for
the IPM forecasts, predicted percent changes in median equilibrium cover
ranged from -200\% to 200\% (Figure 1). Forecast uncertainty did not
vary appreciably among species.

The population responses of species as forecast by the IPM is a function
of the combined effects of individual vital rate sensitivities to
climate and the contribution of each vital to population dynamics. For
\emph{B. gracilis}, all vital rates (survival, growth, and rectruitment)
were positively affected by increases in precipitation and temperature
(Figure 2). Intuitively, this leads to an overall increase in \emph{B.
gracilis} cover (Table 2).

Precipitation increase resulted in all vital rates contributing to an
increase in simulated cover for \emph{H. comata}. But there is a small
negative effect of temperature on recruitment. THIS IS WEIRD!!! CHECK
SIMULATIONS.

\emph{P smithii} responded negatively to an increase in precipitation,
temperature, or both regardless of the vital rate through which
perturbed climate was acting. The effects were small, however, leading
to relatively minor population consequences under climate change that
did not vary much across scenarios (Table 2).

For \emph{B. gracilis}, the IPM predicted a modest increase in cover
with a 1\% increase in the mean of precipitation or a 1\% increase in
the mean of temperature, the compounding effect of both being a 20\%
increase in cover. This reflects the relatively strong effects of
precipitation and climate on \emph{B. gracilis} genet growth and
recruitment (Figure 1). The QBM also predicted increased \emph{B.
gracilis} cover with a precipitation increase, but increasing
temperature decreased equilibrium cover (Figure 3).

@@

\subsubsection{Forecasting the future, and the future of
forecasting}\label{forecasting-the-future-and-the-future-of-forecasting}

Our goal was not make any explicit forecast for the future state of
these populations based on predicted climate change. But our results
highlight the state of affairs in ecology when it comes to forecasting
the impacts of climate change. The analysis we conducted here could be
considered, with some exceptions of course, at the forefront of
ecological forecasting in terms of the statistical approach employed
(hierarchical Bayesian), the type of population model we used
(stochastic IPM with parameter uncertainty), and the amount of data we
had at our disposal (14 years of individual-level data).
